A Logical Solution of the Linearity Challenge

### 2025-12-04 paper/Reduction.lagda.tex
revised description of reduction rules

\begin{AgdaAlign}
\begin{code}
data _↝_ {Γ} : Process Γ → Process Γ → Set where
\end{code}

@@

The constructor \AgdaInductiveConstructor{r-link} models \RLink.

\begin{code}
  r-link      : ∀{Δ A P} (p : Γ ∋ dual A ⊳ Δ) →
                cut {A} p (link (< > •)) P ↝ ↭process (↭concat p) P
\end{code}

For generality, the rules quantifies over two proofs $d$ and $e$ of the same
duality relation between $A$ and $B$. Even though such proofs would necessarily
be the same, distinguishing them saves us a proof that they are equal whenever
we want to apply this reduction. 
%
Also note that the splitting
$(\AgdaInductiveConstructor{split-l}~(\AgdaInductiveConstructor{split-r}~\AgdaInductiveConstructor{split-e}))$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with \Cref{tab:semantics}, where the link $\Link\x\y$ being reduced is oriented
so that the restricted channel $x$ occurs on the left. 
%
Finally, the process $P\subst\y\x$ resulting from the reduction is subject to a
channel substitution in \Cref{tab:semantics}. In the Agda representation of
processes with nameless channels, the same effect is achieved by the permutation
$\AgdaFunction{\#cons}~p$, which ``renames'' (the position of) the free
occurrence of $x$ within $P$ with (the position of) $y$ encoded in the splitting
$p$.

@@

While the \AgdaInductiveConstructor{close} constructor implicitly refers to the
only free channel occurring in a process of the form $\Close\x$, the
\AgdaInductiveConstructor{wait} constructor uses a splitting proof of the form
$(\AgdaInductiveConstructor{split-l}~q_0)$ to be sure that the referenced
channel is also the restricted one, and therefore matches the one of the
\AgdaInductiveConstructor{close} process.
%
Just like \AgdaInductiveConstructor{r-link} quantifies over multiple proofs for
the same duality relation, also \AgdaInductiveConstructor{r-close} (and several
other reduction constructors) quantifies over $p_0$ and $q_0$ which both prove
the splitting $\Context \simeq \AgdaInductiveConstructor{[]} + \Context$. Since
the left context is empty, these splittings must be equal and made of a sequence
of \AgdaInductiveConstructor{split-r} applications followed by
\AgdaInductiveConstructor{split-e}, but it is easier to quantify them separately
so that we do not have to worry about proving their equality whenever we wish to
apply this reduction.

The constructor \AgdaInductiveConstructor{r-select-l} models \RSelect when the
selected tag is $\InTag_1$, that is \AgdaInductiveConstructor{true} in Agda:

@@

There is not much more to note here (and in other omitted reduction rules),
except again for the multiple quantifications over the trivial splittings
$\Context_i \simeq \AgdaInductiveConstructor{[]} + \Context_i$ and the use of
\AgdaInductiveConstructor{split-l} to make sure that the channel referred to by
\AgdaInductiveConstructor{select} and \AgdaInductiveConstructor{case} is indeed
the one restricted by the cut.

\begin{code}[hide]
  r-right     : ∀{Γ₁ Γ₂ A B P Q R}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                cut {A ⊕ B} p (right (< p₀) P)
                              (case (< q₀) Q R) ↝ cut p P R
  r-fork      : ∀{Γ₁ Γ₂ Γ₃ Δ A B P Q R}
                (p : Γ ≃ Δ + Γ₃) (p₀ : Γ₃ ≃ [] + Γ₃) (q : Δ ≃ Γ₁ + Γ₂) (q₀ : Δ ≃ [] + Δ) →
                let _ , p′ , q′ = +-assoc-l p q in
                cut {A ⊗ B} p (fork (< q₀) q P Q)
                              (join (< p₀) R) ↝ cut q′ P (cut (> p′) Q R)
  r-client    : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
                cut {`! A} p (server (< p₀) un P) (client (< q₀) Q) ↝ cut p P Q
  r-weaken    : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
                cut {`! A} p (server (< p₀) un P)
                             (weaken (< q₀) Q) ↝ weakening un p Q
  r-contract  : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
                cut {`! A} p (server (< p₀) un P)
                             (contract (< q₀) Q) ↝
                    contraction un p
                      (cut ++≃+ (server (< p₀) un P)
                      (cut (> p) (server (< p₀) un P) Q))
  r-poly       : ∀{A B Γ₁ Γ₂ P F} (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) ->
                 cut {`∃ A} p (ex {_} {B} (< p₀) P) (all (< q₀) F) ↝ cut p P (F B)
\end{code}

@@

\begin{code}
  r-cut        : ∀{Γ₁ Γ₂ A P Q R} (q : Γ ≃ Γ₁ + Γ₂) →
                 P ↝ Q → cut {A} q P R ↝ cut q Q R
  r-cong       : ∀{P R Q} → P ⊒ R → R ↝ Q → P ↝ Q
\end{code}
\end{AgdaAlign}

>>>>

\begin{code}
data _↝_ {Γ} : Process Γ → Process Γ → Set where
\end{code}

@@

The constructor \AgdaInductiveConstructor{r-link} models the reduction
$\Cut[A]\x{\Link\x\y}{P} \red P\subst\y\x$ called \RLink in
\Cref{tab:semantics}:

\begin{code}
  r-link      : ∀{Δ A P} (p : Γ ≃ [ dual A ] + Δ) →
                cut {A} p (link (< > •)) P ↝ ↭process (↭concat p) P
\end{code}

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ occurs on the left. 
%
The process $P$ has type
$\AgdaDatatype{Process}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is found in the splitting
$p$. We ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

@@

While the \AgdaInductiveConstructor{close} constructor implicitly refers to the
only free channel occurring in a process of the form $\Close\x$, the
\AgdaInductiveConstructor{wait} constructor uses a splitting proof of the form
$(\AgdaInductiveConstructor{<}~q_0)$ to be sure that the referenced channel is
also the restricted one, and therefore matches the one of the
\AgdaInductiveConstructor{close} process.

Note that \AgdaInductiveConstructor{r-close} (and several other reduction
constructors) quantifies over $p_0$ and $q_0$ which both prove the splitting
$\Context \simeq \AgdaInductiveConstructor{[]} + \Context$. Since the left
context is empty, these splittings must be equal and made of a sequence of
\AgdaInductiveConstructor{>} applications followed by
\AgdaInductiveConstructor{•}. In general, Agda will not be able to ``see'' that
they are definitionally equal, hence it is easier to quantify them separately so
that we do not have to worry about proving their equality whenever we wish to
apply this reduction.

The constructor \AgdaInductiveConstructor{r-left} models \RSelect when the
selected tag is $\InTag_1$:

@@

There is not much more to note here except again for the multiple
quantifications over the trivial splittings $\Context_i \simeq
\AgdaInductiveConstructor{[]} + \Context_i$ and the use of
\AgdaInductiveConstructor{<} to make sure that the channel referred to by
\AgdaInductiveConstructor{left} and \AgdaInductiveConstructor{case} is indeed
the one restricted by the cut.

\begin{code}[hide]
  r-right     : ∀{Γ₁ Γ₂ A B P Q R}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                cut {A ⊕ B} p (right (< p₀) P)
                              (case (< q₀) Q R) ↝ cut p P R
  r-fork      : ∀{Γ₁ Γ₂ Γ₃ Δ A B P Q R}
                (p : Γ ≃ Δ + Γ₃) (p₀ : Γ₃ ≃ [] + Γ₃) (q : Δ ≃ Γ₁ + Γ₂) (q₀ : Δ ≃ [] + Δ) →
                let _ , p′ , q′ = +-assoc-l p q in
                cut {A ⊗ B} p (fork (< q₀) q P Q)
                              (join (< p₀) R) ↝ cut q′ P (cut (> p′) Q R)
  r-client    : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
                cut {`! A} p (server (< p₀) un P) (client (< q₀) Q) ↝ cut p P Q
  r-weaken    : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
                cut {`! A} p (server (< p₀) un P)
                             (weaken (< q₀) Q) ↝ weakening un p Q
  r-contract  : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
                cut {`! A} p (server (< p₀) un P)
                             (contract (< q₀) Q) ↝
                    contraction un p
                      (cut ++≃+ (server (< p₀) un P)
                      (cut (> p) (server (< p₀) un P) Q))
\end{code}

The remaining constructors that describe the base reductions follow a similar
pattern. We only glance at \RExists since it involves a non-trivial rewriting of
types:

\begin{code}
  r-ex         : ∀{A B Γ₁ Γ₂ P F}
                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                 cut {`∃ A} p (ex {_} {B} (< p₀) P) (all (< q₀) F) ↝ cut p P (F B)
\end{code}

Recalling the definitions of \AgdaInductiveConstructor{ex} and
\AgdaInductiveConstructor{all} from \Cref{sec:process-agda}, we note that $P$
has type
$\AgdaDatatype{Process}~(\AgdaFunction{subst}~[~B~/\_]~A~\AgdaInductiveConstructor{∷}~\Context_1)$
and $(F~B)$ is a process of type
$\AgdaDatatype{Process}~(\AgdaFunction{subst}~[~B~/\_]~(\AgdaFunction{dual}~A)~\AgdaInductiveConstructor{∷}~\Context_2)$.
In order for these two process to be composable in a cut, it must be the case
that $\AgdaFunction{subst}~[~B~/\_]~(\AgdaFunction{dual}~A) \equiv
\AgdaFunction{dual}~(\AgdaFunction{subst}~[~B~/\_]~A)$, which was proved in
\Cref{sec:type-agda} under the name \AgdaFunction{dual-subst}. Thanks to the
implicit rewriting rule, we do not have to rewrite the index in the type of
$(F~B)$, which is silently accepted as is.

@@

\begin{code}
  r-cut        : ∀{Γ₁ Γ₂ A P Q R} (q : Γ ≃ Γ₁ + Γ₂) →
                 P ↝ Q → cut {A} q P R ↝ cut q Q R
  r-cong       : ∀{P R Q} → P ⊒ R → R ↝ Q → P ↝ Q
\end{code}
